### 2020-08-27 thesis/thesis.tex
[thesis] Extend sections about cohandlers
+ Various minor corrections

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations, which
    in programming langauges, may be a way to model interaction with external
    resources, where based on current configuration and some parameter, we
    obtain new configuration along with it's result.

@@

    Practical part that is closest to topic of the thesis, is work done by
    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
    calculus with runners of coalgebraic effects that ensures linear usage of
    resources as well as execution of code handling finalisation.

@@

\section{Data-Flow Analysis}

@@

Despite of the above means, algebraic effects that occur in the source code may
drop the resumption and thus finalization code is not going to be executed.
We address this issue by not allowing effects to escape the cohandler, which
means that we are not going to discard the execution of cohandler.

@@

@@

    % Outdated
    As of this day, two implementations are available, one based on the curried
    translation from Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit
    stacks~\cite{handlers-cps-journal}. More details can be found in
    Section~\ref{chapter:implementation}. All commands are available within the
    \verb!src! directory.

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
        \item Compile: \verb!make build!
        \item Link to PATH:~\verb!sudo make link!
        \item Remove artifacts: \verb!make clean!
    \end{itemize}

@@

    \begin{itemize}
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
        \item Compile, run linter and tests: \verb!make check!
    \end{itemize}

@@

    Semantic of Freak is heavily based on Links langauge~\cite{handlers-cps-journal},
    for which the source language's dynamics have been described extensively by
    providing small-step operational semantics, continuation passing style
    transformation~\cite{handlers-cps} as well as abstract
    machine~\cite{liberating-effects}, which was proved to coincide with CPS
    translation. That being said, Freak introduces new basic constructs to the
    language, for which we shall define the semantics.

@@

    \subsection{Cohandlers}

    TODO cohandlers

@@

\begin{verbatim}
    data Value
        = VVar Var
        | VNum Integer
        | VStr String
        | VLambda Var ValueType Comp
        | VFix Var Var Comp
        | VUnit
        | VPair Value Value
        | VRecordRow (RecordRow Value)
        | VExtendRow Label Value Value
        | VVariantRow (VariantRow Value)
        | VBinOp BinaryOp Value Value

    data Comp
        = EVal Value
        | ELet Var Comp Comp
        | EApp Value Value
        | ESplit Label Var Var Value Comp
        | ECase Value Label Var Comp Var Comp
        | EReturn Value
        | EAbsurd Value
        | EIf Value Comp Comp
        -- Algebraic effects
        | EOp Label Value
        | EHandle Comp Handler
        -- Coalgebraic effects
        | ECoop Label Value
        | ECohandle Comp Handler
\end{verbatim}

    \noindent
    Similarly for the target calculus data structure. However, as one may notice,
    for convenience the \textbf{let} translation is homomorphic, as opposed to be
    to lambda abstracted with immediate application:

\begin{verbatim}
    data UValue
        = UVar Var
        | UNum Integer
        | UStr String
        | UBool Bool
        | ULambda Var UComp
        | UUnit
        | UPair UValue UValue
        | ULabel Label
        | URec Var Var UComp
        | UBinOp BinaryOp UValue UValue

    data UComp
        = UVal UValue
        | UApp UComp UComp
        | USplit Label Var Var UValue UComp
        | UCase UValue Label UComp Var UComp
        | UIf UValue UComp UComp
        | ULet Var UComp UComp
        | UAbsurd UValue
        | UTopLevelEffect Label UValue
\end{verbatim}

    \noindent
    The final answer, common to both evaluations, is represented as a \verb!DValue!,
    where the meaning of the coproduct is as one would expect:

\begin{verbatim}
    type Label = String

    type FuncRecord = [DValue] -> Either Error DValue

@@

    \noindent
    Evaluation of target calculus is typical to call-by-value with syntactic
    distinction between values and computations. For more details, refer
    to~\cite{handlers-cps-journal}. That being said, there is one part
    that requires more attention, namely, \verb!UTopLevelEffect Label UValue!,
    which represents an unhandled algebraic effect along with it's parameter.
    Even in case language has coalgebraic effects, special treatment is required
    for doing IO\@. For that reason, we define default handling for a few effects
    for printing to console and handling files, where semantic is just as one
    would expect from naming. We abuse haskell's notation to define types for
    algebraic effects living in Freak langauge:

@@

    initialEffCont :: ContF
    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
        return . UApp (UVal r) $ UTopLevelEffect effLabel p

@@

\begin{verbatim}
cps     :: Comp    -> [Cont] -> CPSMonad UComp
cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
cpsOp   :: Label -> Value -> [Cont] -> CPSMonad UComp
cpsHRet :: Handler -> Cont
cpsHOps :: (ContF -> Cont) -> Handler -> Cont
forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
runCPS  :: Comp -> EvalResMonad UComp
\end{verbatim}

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation, where evaluation result is represented by\\
    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
    only deep ones.

@@

\begin{verbatim}
    AST.hs          - AST data structures
    CommonCPS.hs    - Common functions for CPS translation
    CommonEval.hs   - Common functions for evaluation
    CPSLists.hs     - Uncurried CPS translation
    EvalTarget.hs   - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    TargetAST.hs    - AST for the target calculus
    Tests.hs        - Tests module
    Types.hs        - Common types definition
    programs/       - Exemplary programs covered in tests

@@

    TODO What was achieved, what was described, what needs to be done or researched.

@@

    The Links language, on which CPS translation Freak was based, also provides
    small-step operational semantics~\cite{handlers-cps} and an abstract
    machine~\cite{liberating-effects}. Implementing another way of evaluation
    could serve as a way to empirically assert correctness, as opposed to formally.
    Abstract machine would need to be extended to incorporate coeffects.

    \subsection{Type inference and row polymorphism}

    The type system as of this day is not implemented, as the focus has been put
    on CPS transformation and dynamics of the calculus with effects and coeffects.
    Further work is required here, especially considering the fact that a huge
    advantage of algebraic effects is that they are explicitly defined in the
    type of a computation.

@@

    The Freak language is limited to a single instance of an effect. We would
    need to support cases where many instances of the algebraic effects, with
    the same handler code, could be instantiated. The current state of the
    art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
    or instance variables, as in Helium~\cite{binders-labels}. Another part is
    to research on how this area combines with coalgebraic part.

@@

    Exceptions are a trivial example of algebraic effect where the resumption is
    discarded, and as described in \S 4.5~\cite{handlers-cps}, they can be modeled
    as a separate construct to improve performance.

@@

    Shallow and deep handlers while being able to simulate each other up to
    administrative reductions, have a very different meaning from a theoretical point
    of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
    could be another way of enhancing Freak.

>>>>

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations, which
    in programming languages, may be a way to model interaction with external
    resources, where based on current configuration and some parameter, we
    obtain new configuration along with it's result.

@@

    Practical part that is closest to topic of the thesis, is work done by
    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
    implementation of a $\lambda_{coop}$ language~\cite{coop} and
    $\mathVar{haskell-coop}$~\cite{haskell-coop} library, which is a
    calculus with runners of coalgebraic effects that ensures linear usage of
    resources as well as execution of code handling finalisation.

@@

\section{Data-Flow Analysis}\label{sec:data-flow}

@@

Despite of the above means, algebraic effects that occur in the source code may
drop the resumption and thus finalization code is not going to be executed.
We address this issue by not allowing effects to escape the cohandler, which
means that we are not going to allow to discard the execution of cohandler.

@@

Full and working example with implementation of automata can be found under
\verb!programs/coeffects/NFA.fk! in Freak repository~\cite{freak}.

@@

    As of this day, two implementations are available, one based on the curried
    translation from Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit
    stacks~\cite{handlers-cps-journal}. More details can be found in
    Section~\ref{chapter:implementation}. All commands are available within the
    \verb!src! directory. Please note, however, that Apple-based version is no
    longer supported, and is left as a reference.

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Select implementation: \verb!make link-lists! (recommended) vs \verb!make link-appel!
        \item Compile: \verb!make build!
        \item Link to PATH:~\verb!sudo make link!
        \item Remove artifacts: \verb!make clean!
    \end{itemize}

@@

    \begin{itemize}
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
        \item Validate Freak code from \verb!programs! dir: \verb!make lint!
        \item Run all above commands: \verb!make check!
    \end{itemize}

@@

    Semantic of Freak is heavily based on Links language~\cite{handlers-cps-journal},
    for which the source language's dynamics have been described extensively by
    providing small-step operational semantics, continuation passing style
    transformation~\cite{handlers-cps} as well as abstract
    machine~\cite{liberating-effects}, which was proved to coincide with CPS
    translation. That being said, Freak introduces new basic constructs to the
    language, for which we shall define the semantics.

@@

    \subsection{Cohandlers}\label{subsec:calc-cohandlers}

    Semantics for cohandlers is brought from denotational one that we have
    introduced in Section~\ref{sec:duality}, meaning, we denote cooperations
    as state passing transitions in a code, represented similarly as algebraic
    effects, through (co)free (co)models of algebraic theories along with their
    morphisms. Let's put the following cooperation:

    \begin{center}
        $ {\llbracket op \rrbracket}^W : {|C|} \times {P} \rightarrow  {|C|} \times {X} $
    \end{center}

    Operationally, we embed transitions between coeffectull computations
    in a standard state monad $ S_{C}X \triangleq (C \Rightarrow X \times C) $.
    We achieve this by transforming AST of the source code, such that
    configurations are passed among cooperations. For a given coalgebraic theory
    $T$, handler of the state monad in Freak looks as follows:

\vbox{\begin{verbatim}
{
    TPut s' r -> return (\s -> let g <- r () in g s') |
    TGet _ r -> return (\s -> let g <- r s in g s) |
    return x -> return (\s -> return x)
}
\end{verbatim}
}

    \noindent
    Having that defined, we expand computation

\begin{verbatim}
    observe Coeffect p
\end{verbatim}

    \noindent
    into the following one (modulo $\mathVar{alpha-conversion}$):

\vbox{\begin{verbatim}
    let Tconf <- observe TGet () in
    let Tcoresult <- observe Coeffect (Tconf, p) in
    let TnewConf <- return fst Tcoresult in
    let Tres <- return snd Tcoresult in
    let _ <- observe TPut TnewConf in
    return Tres
\end{verbatim}
}

    \noindent
    where before executing a coeffect, we obtain current configuration,
    pass it to coeffect, put new config and then return the actual result.
    This is done behind the scenes and user does not know about state
    passing.

@@

\begin{verbatim}
    data EffLabel = EffL String
                  | CoeffL String

    data Value
        = VVar Var
        | VNum Integer
        | VStr String
        | VLambda Var ValueType Comp
        | VFix Var Var Comp
        | VUnit
        | VPair Value Value
        | VRecordRow (RecordRow Value)
        | VExtendRow Label Value Value
        | VVariantRow (VariantRow Value)
        | VBinOp BinaryOp Value Value

    data Comp
        = EVal Value
        | ELet Var Comp Comp
        | EApp Value Value
        | ESplit Label Var Var Value Comp
        | ECase Value Label Var Comp Var Comp
        | EReturn Value
        | EAbsurd Value
        | EIf Value Comp Comp
        -- Algebraic effects
        | EOp EffLabel Value
        | EHandle Comp Handler
        -- Coalgebraic effects
        | ECoop EffLabel Value
        | ECohandle Comp Handler
        -- Intermediate representation for a cohandler
        | ECohandleIR AlgTheoryName Value Comp Handler

    data Handler
        = HRet Var Comp
        | HOps AlgebraicOp Handler

    data AlgebraicOp = AlgOp EffLabel Var Var Comp
\end{verbatim}

    \noindent
    Where under \verb!Value! in \verb!ECohandleIR! we store initial
    configuration of a coalgebra, and under \verb!AlgTheoryName! name of the
    theory over which we operate. We use \verb!EffLabel! structure to
    syntactically distinguish between effects and coeffects in the code.

    Data structures for the calculus are also driven by syntax. However, as one
    may notice, for convenience the \textbf{let} translation is homomorphic,
    as opposed to be to lambda abstracted with immediate application:

\begin{verbatim}
    data UValue
        = UVar Var
        | UNum Integer
        | UStr String
        | UBool Bool
        | ULambda Var ([Cont] -> CPSMonad UComp)
        | UUnit
        | UPair UValue UValue
        | ULabel Label
        | UEffLabel EffLabel
        | URec Var Var UComp
        | UBinOp BinaryOp UValue UValue

    data UComp
        = UVal UValue
        | UApp UComp UComp [Cont]
        | USplit Label Var Var UValue UComp
        | UCase UValue Label UComp Var UComp
        | UIf UValue UComp UComp
        | ULet Var UComp UComp
        | UAbsurd UValue
        | UTopLevelEffect EffLabel UValue
\end{verbatim}

    \noindent
    Careful reader may notice the correspondence between \verb!UApp! having
    a stack of continuations, and \verb!ULambda! being a
    \verb![Cont] -> CPSMonad UComp! function, which follows directly from
    translation proposed by Lindley et al\.~\cite{handlers-cps-journal}.

    The final answer, common to both evaluations, is represented as a \verb!DValue!,
    where the meaning of the coproduct is as one would expect. At this point,
    we no longer treat effect labels as a special case.

\begin{verbatim}
    type Label = String
    type EvalMonad a = ExceptT Error (StateT Int IO) a
    type FuncRecord = [DValue] -> [Cont] -> EvalMonad DValue

@@

    Evaluation of target calculus is typical to call-by-value with syntactic
    distinction between values and computations. For more details, refer
    to~\cite{handlers-cps-journal}. That being said, there is one part
    that requires more attention, namely, \verb!UTopLevelEffect Label UValue!,
    which represents an unhandled algebraic effect along with it's parameter.
    Even in case language has coalgebraic effects, special treatment is required
    for doing IO\@. For that reason, we define default handling for a few effects
    for printing to console and handling files, where semantic is just as one
    would expect from naming. We abuse haskell's notation to define types for
    algebraic effects living in Freak language:

@@

    initialEffCont :: ContF
    initialEffCont (UPair (UEffLabel l) (UPair p r)) ks =
        return $ UApp (UVal r) (UTopLevelEffect l p) k

@@

\begin{verbatim}
data EffT = EffT | CoeffT
type EvalResMonad a = IO (Either Error a)

cps     :: Comp     -> [Cont]  -> CPSMonad UComp
cpsVal  :: Value    -> [Cont]  -> CPSMonad UValue
cpsOp   :: EffLabel -> Value   -> [Cont] -> CPSMonad UComp
cpsHRet :: Handler  -> Cont
cpsHOps :: EffT     -> Handler -> Cont
forward :: EffLabel -> UValue  -> UValue -> [Cont] -> CPSMonad UComp
runCPS  :: Comp     -> EvalResMonad UComp
\end{verbatim}

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our language implements
    only deep ones.

@@

\begin{verbatim}
    AST.hs          - AST data structures
    CommonCPS.hs    - Common functions for CPS translation
    CommonEval.hs   - Common functions for evaluation
    CPSLists.hs     - Uncurried CPS translation
    EvalTarget.hs   - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    TargetAST.hs    - AST for the target calculus
    Tests.hs        - Tests module
    Types.hs        - Common types definition
    Transform.hs    - Cohandlers transformation described in Calculus chapter
    programs/       - Exemplary programs covered in tests

@@

    We have showed derivation of cooperations, based on previous work, as well
    as presented a calculus combining algebraic effects with their dual notion.
    This resulted in a programming language which allows user to seemlessly
    operate on resources based on solid mathematical foundations. It can seen now,
    that (co)algebraic effects are an abstract concept that captures plenty of
    patterns in everyday programming, which already had ad-hoc and more
    restricted implementations in many existing languages, through constructs
    such as \verb!try, except, finally!, or by more low-level mechanisms
    such as handling interrupts in operating systems.

    This relatively new concept in theory of progamming languages yet needs to
    be researched, and we propose in the next section a few directions into
    which Freak language could go.

@@

    The Links language, on which CPS translation Freak was based, also provides
    small-step operational semantics~\cite{handlers-cps} and an abstract
    machine~\cite{liberating-effects}. Implementing another way of evaluation
    could serve as a way to empirically assert correctness, as opposed to formally.
    Coeffects are living in a syntactic world, which are then translated to
    algebraic effects, thus we may not need to extend the abstract machine to
    accomodate them.

    \subsection{Static analysis}

    The type system as of this day is not implemented, as the focus has been put
    on CPS transformation and dynamics of the calculus with effects and coeffects.
    Further work is required here, especially considering the fact that a huge
    advantage of algebraic effects is that they are explicitly defined in the
    type of a computation. We have been considering row polymorphism as the
    tool of choice.

    Another part could involve statics through data-flow analysis, which we
    have already mentioned in Section~\ref{sec:data-flow}.

@@

    Languages with algebraic effects tend to have an mechanism, in which one
    can instantiate multiple versions of the effects under the same handler
    code. The current state of the art introduces a concept of resources and
    instances, as in Eff~\cite{programming-in-eff}, or instance variables, as
    in Helium~\cite{binders-labels}. Work in this area could involve diving
    into how this concept is related to coalgebras, and whether to implement
    that in Freak.

@@

    Exceptions are a trivial example of algebraic effect where the resumption is
    discarded, and as described in \S 4.5~\cite{handlers-cps} they can be
    modeled as a separate construct. On the one hand, it may lead to
    performance improvements, on the other, we could use this notion to ensure
    that finalisation code for coeffects is executed even in the case of a
    failure, as described in~\cite{runners-in-action}.

@@

    Shallow and deep handlers while being able to simulate each other up to
    administrative reductions, have a very different meaning from a theoretical point
    of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
    could be another way of enhancing the language.
